\bf Title here

### 2020-06-09 paper/paper.tex
more text editing - whole paper except 1.3 & 1.4

\bigskip
\begin{abstract}
\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  check consistency of using PP for projection pursuit and PPI for
  projection pursuit index
\end{enumerate}
\end{abstract}

@@

Given an objective function \(f\), one way of optimising it is to equate
its gradient to zero. In modern optimisation problems, gradient
information can be hard to evaluate or sometimes even impossible and
Derivative-Free Optimisation (DFO) methods can be useful to approach
these problems. The two most common methods in DFO are
\emph{direct-search methods} and \emph{model-based methods} and this
paper dedicates to the discussion of direct-search methods, which gains
its popularity due to its simplicity in use and reliability in
complicated practical problem. Coined by \citet{hooke1961direct}, direct
search methods don't require any gradient or hessian information and
evaluate \(f\) directly. A well-known example of it is the Nelder-Mead
algorithm \citep{nelder1965simplex} and it enjoys the popularity due to
its simplicity and reliability. {[}feel like this sentence can be
expanded to include more information. xxx{]}. Further, direct search
methods can be classified as \emph{stochastic} and \emph{deterministic}
depends on whether a random sample is used in the search. The stochastic
version of direct-search method is the focus of this paper.

@@

The optimisation problem we're interested in is in the context of
projection pursuit. Coined by \citet{friedman1974projection}, projection
pursuit is a method that detects the interesting structure
(i.e.~clustering, outliers and skewness) of multivariate data via
projecting it in lower dimensions. A Projection Pursuit Index (PPI) is a
function of the projection matrix (a.k.a projection angle, projection
basis) measuring the ``interestingness'' of data and we refer the value
of PPI function as \emph{index value}. In the literature, the indices
being proposed include lengendre index \citep{friedman1974projection},
hermite index \citep{hall1989polynomial}, natural hermite index
\citep{cook1993projection}, chi-square index
\citep{posse1995projection}, LDA index \citep{lee2005projection} and PDA
index \citep{lee2010projection}. {[}Any literature on holes index?
xxx{]}

@@

Projection pursuit is usually used in conjunction with a tour method
called \emph{guided tour}. Tour explores the multivariate data
\emph{interactively} via playing a series of projections, that form a
\emph{tour path}, like movie frames in real time. Starting from a
randomly generated projection basis, projection pursuit searches for a
target basis with higher index value and guided tour interpolates
geodesically between the initial basis and the target basis. This series
of bases forms a tour path and the projection of the data on these bases
is played in real time. This search-and-interpolateprocess is repeated
until no basis can be found with higher index value. Modified from
\citet{buja2005computational}, Figure \ref{tour-path} vividly depicts
the tour path in guided tour. Guided tour, along with other types of
tour, has been implemented in the \emph{tourr} package in R, available
on the Comprehensive R Archive Network at
\url{https://cran.r-project.org/web/packages/tourr/}
\citep{wickham2011tourrpackage}.

@@

\begin{itemize}
\item
  \emph{Being able to handle non-differentiable PPI function}: The PPI
  function could be non-differentiable, thus derivative free methods are
  preferred.
\item
  \emph{Being able to optimise with constraints}: The constraint comes
  from projection matrix being an orthonormal matrix.
\item
  \emph{Being able to find both local and global maximum}: Although the
  primary interest is to find the global maximum, local maximum could
  also reeal structures in the data that are of our interest.
\end{itemize}

@@

Random search methods has a black-box machinism and focuses solely on
finding the global maximum point, while the projection pursuit problem
we have is aiming at \emph{exploring} the data and thus is interested in
how the algorithm finds its maximum. This motivates us \textbf{to
develop a visual diagnostic system for exploring the optimisation
searching path}.

The necessity of developing such a system rather than simply producing
different diagnostic plots is because there is always needs to develop a
function specific for diagnosing one particular variables and these
functions developed are not reusable at all. Thus we want to establish a
set of rules that prescribe the diagnostic of iterative algorithms
systematically.

The idea of unifying all the diagnostic plots under one framework is
inspired by the concept of \emph{gramma of
graphic}\citep{wickham2010layered}, which powers the primary graphical
system in R, ggplot2 \citep{ggplot2}. In gramma of graphic, plots are
not defined by its appearance (i.e.~boxplot, histogram, scatter plot
etc) but by ``stacked layers''. By this design, ggplot doesn't need to
develop a gazillion of functions that each produces a different type of
plot, instead, it defines layers and allow users to specify the plot
through these defined layers.

{[}Do I need this paragraph to details the gramma of graphic? xxx{]}

\hypertarget{glocal-object}{%
\subsection{Glocal object}\label{glocal-object}}

Ggplot prescribes how to make a graphic but it doesn't prescribe how to
run a diagnostic. {[}need to have a paragraph or two here to describe
how I come up with the design of four components in the global object -
need to related to the component in optimisation. xxx{]}.

A global object used for diagnosing an optimisation is defined with four
components: decision variable, value of objective function, iterators
and other parameter of interest. In the projection pursuit guided tour,
the decision variable is the projection matrix, the value of objective
function is index value. There are three iterators: \texttt{id} is the
smallest ordering unit that increases by one for each observatoin;
\texttt{tries} is a larger order unit, which is updated once a
search-and-interpolate step is finished. A third iterator \texttt{loop}
is used to record the number of repitition in the search step and starts
over from one at the beginning of a new \texttt{tries}. Since there are
different methods used in the projection pursuit guided tour, we are
interested to compare between different methods, thus \texttt{method} is
created as another parameter of interest. The neighbourhood parameter
alpha is always interested in the direct search methods since it
controls the size of the sampling space, thus \texttt{alpha} is another
parameter of interest. \texttt{info} is created to label different stage
in the searching process and it is the third parameter of interest. A
sketch of the global object for projection pursuit guided tour is
presented in Figure \ref{fig:glb-obj}.

@@

Two set of simulated data are used in the demo of the visual diagnostics
in projection pursuit guided tour. A small dataset consists of 1000
randomly simulated observations of five variables (\texttt{x1},
\texttt{x2}, \texttt{x8}, \texttt{x9}, \texttt{x10}). \texttt{x2} is the
informative variable simulated from two bi-modal normal distribution
centred at -3 and 3 with variance of one and the other four are
simulated from non-informative standard random normal distributions. The
data has been scaled to ensure \texttt{x2} has variance of 1. The goal
is to find the 1D projection of bi-modal shape, which correspondes to
the projection matrix (vector) of \texttt{{[}0,\ 1,\ 0,\ 0,\ 0{]}}.

A larger dataset contains more informative variables (\texttt{x3} to
\texttt{x7}) of different types. The distribution of all the variables
except \texttt{x3} is plotted below and \texttt{x3} takes 500 positive
one and 500 negative one. {[}should I introduce the dist for each var?
xxx{]} {[}also feel like we don't really use \texttt{x3} to \texttt{x6},
should we not mention about these? xxx{]}

\includegraphics{paper_files/figure-latex/origin-data-1.pdf}

Once the dataset is sent to the tourr package, all the information
generated will be stored in a global structure. The global structure
consists of six columns: \texttt{basis}, \texttt{index\_val},
\texttt{tries}, \texttt{info}, \texttt{loop}, \texttt{id} and captured
all the basis generated during whole tour process. The example below
presents the global object of a 1D projection of the small dataset with
geodesic searching method.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{holes_1d_geo }\OperatorTok{%>%}\StringTok{ }\KeywordTok{head}\NormalTok{(}\DecValTok{5}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

\texttt{tries} has an increment of one once the generator is called
(equivalently a new target basis is generated); \texttt{info} records
the stage the basis is in. This would include the \texttt{interpolation}
stage and the detailed stage in the optimisation i.e.
\texttt{direction\_search}, \texttt{best\_direction\_search},
\texttt{line\_search}and \texttt{best\_line\_search} for geodesic
searching (\texttt{search\_geodesic}); \texttt{random\_search} and
\texttt{new\_basis} for simulating annealing (\texttt{search\_better}).
\texttt{loop} is the counter used for the optimisation procedure and
thus will be \texttt{NA} for interpolation steps. \texttt{id} creates a
sequential order of the basis. This information will be stored and
printed when the optimisation ends and can be turned off via
\texttt{print\ =\ FALSE}. Additional messages during the optimisation
can be displayed via \texttt{verbose\ =\ TRUE}. Another examples is a 2D
projection of the larger dataset with two informative variable
(\texttt{x2} and \texttt{x7}) using search\_better method. Notice in
this example, the dimension of the bases becomes 6 by 2.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{holes_2d_better }\OperatorTok{%>%}\StringTok{ }\KeywordTok{head}\NormalTok{(}\DecValTok{5}\NormalTok{) }
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## # A tibble: 5 x 8
##   basis             index_val tries info           loop method       alpha    id
##   <list>                <dbl> <dbl> <chr>         <dbl> <chr>        <dbl> <int>
## 1 <dbl[,2] [6 x 2]>     0.804     1 start            NA <NA>           0.5     1
## 2 <dbl[,2] [6 x 2]>     0.793     1 random_search     1 search_bett~   0.5     2
## 3 <dbl[,2] [6 x 2]>     0.784     1 random_search     2 search_bett~   0.5     3
## 4 <dbl[,2] [6 x 2]>     0.773     1 random_search     3 search_bett~   0.5     4
## 5 <dbl[,2] [6 x 2]>     0.795     1 random_search     4 search_bett~   0.5     5
\end{verbatim}

@@

The most interesting parameter to explore is the value of objective
function. The points recorded in the global object can be divided into
two broad categories: searching points and interpolating points

\begin{itemize}
\item
  \emph{Searching points} include the observations that are recorded in
  the searching algorithm in order to find the target basis. The points
  for target bases is also included in the searching points and there is
  one such point per \texttt{tries}.
\item
  \emph{interpolating points} exist in the guided tour to produce
  continuous animated view from one target basis to another and it
  doesn't have \texttt{loop} value.
\end{itemize}

@@

As mentioned previously, the largest difficulties of exploring searching
points is its unknown number of observations per \texttt{tries}. Mapping
\texttt{id} on the x-axis will leave the \texttt{tries} with few
observations a small space in the plot, while those \texttt{tries} with
large number of search points towards the end occupying the vast
majority of the space in the plot.

This motivates the use of summarised statistics. At each iteration,
rather than knowing the index value of \emph{every} points, we are more
interested to know a general summary of all the points and more
importantly, the point with the largest \texttt{index\_val} since it
prescribes the geodesic interpolation and future searches.

Boxplot is a suitable candidate that provides five points summary of the
data, while it has one drawback: it doesn't report the number of point
in each box. We may risk losing information on how long it takes for the
search to find the target basis by displaying the boxplot alone for all
\texttt{tries}. Thus, the number of point for each \texttt{tries} is
displayed at the bottom of each box and we provide options to switch
\texttt{tries} with small number of points to a point geometry. This is
achieved via the \texttt{cutoff} argument. A line geometry is also added
to link the points with the largest index value for each \texttt{tries}.
This helps to visualise the improvement made by each \texttt{tries}.

@@

\emph{Example: exploring searching points} The data is sourced from a 2D
projection of the larger dataset and \texttt{search\_better} is used
with \texttt{max.tries\ =\ 500}. In Figure \ref{fig:points-id} and
\ref{fig:points}, a comparison has been made on visualising the
searching points with \texttt{id} and \texttt{tries} on the x-axis,
colored by \texttt{tries}. In Figure \ref{fig:points-id}, the searching
points of the first few \texttt{tries} are squeezed in a small width
leaving the large uninteresting searching points in the last three
\texttt{tries} taking a vast majority of the plotting space. While in
Figure \ref{fig:points} the data is spaced by \texttt{tries} evenly in
the plot. Label at the bottom indicates the number of observations in
each tries and facilitates the choice of cutoff to switch from point
geometry to boxplot geometry (\texttt{cutoff\ =\ 15}). The line geometry
suggest the largest improvement happens at \texttt{tries\ =\ 5}.

@@

Plotting the interpolating points as time series data allows us to
diagnose characteristics of different configurations and index
functions. Here we present two examples of using plots to diagnose the
tour algorithm and different index functions.

\emph{Example: Interruption} This examples uses \texttt{search\_better}
for a 2D projection on the larger dataset using the \texttt{holes}
index. As mentioned previously, the interpolation starts from the
current basis to the target basis, which has been found by the
projection pursuit algorithm to have a higher index value. After the
interpolation, the target basis will become the current basis and send
back to the projection pursuit algorithm to find the next target basis.
From figure \ref{fig:interruption}, it is possible that there are bases
with index value higher than the target basis on the interpolation path
and these bases could be used to search for new basis in the next
iteration.

Thus an interruption is constructed to accept the interpolating bases up
to the one with the larger index value on the interpolation path, and
that basis is taken as the current basis for the next iteration. After
implementing this interruption, the tracing plot with the same
configuration is shown on the lower panel. We can observe that rather
than interpolating to the target basis at \texttt{id\ =\ 62}, the
interpolation stops before the index value starts to decrease at
\texttt{id\ =\ 60}. This implementation results in a higher index value
in the end with fewer steps.

@@

\emph{Example: Noisy index function} The interpolation path of holes
index, as seen in Figure \ref{fig:interruption}, is smooth, while this
may not be the case for more complicated index function.
\texttt{kol\_cdf} index, an index function based on Kolmogorov test,
compares the difference between a projection matrix and a randomly
generated normal distribution based on cumulated distribution function
(CDF). Several visualisations below shows some characteristic of this
index function.

Figure \ref{fig:geodesic-better} compares the tracing plot of the
interpolating points for \texttt{search\_geodesic} and
\texttt{search\_better}. Rather than a smooth interpolation, the
interpolation path for \texttt{kol\_cdf} shows a zig-zag pattern.
Polishing step has done much more work to reach the final index value
for \texttt{search\_geodesic} than \texttt{search\_better} and this
indicates that noisy index's favour of a random search method than
ascent method.

\begin{figure}
\centering
\includegraphics{paper_files/figure-latex/geodesic-better-1.pdf}
\caption{Comparison of two different searching methods: search\_geodesic
and search\_better on 1D projection problem for a noisy index: kol\_cdf.
The geodesic search rely heavily on the polishing step to find the final
index value while search better works well.}
\end{figure}

Moving from a 1D projection on a dataset with only one informative
variable, we now introduce another informative variable in the data and
remain 1D projection. In this problem, there are two local maximum being
0.17 when the projection matrix is \([0, 1, 0, 0, 0, 0]\) and 0.228 when
the projection matrix is \([0, 0, 1 ,0, 0, 0]\). As in Figure
\ref{fig:1d-2var-different-seeds}, using \texttt{search\_better} with
different seeds, both local maximum can be found. A simulated annealing
algorithm, \texttt{search\_better\_random} in Figure
\ref{fig:1d-2var-better-random}, can always find the global maximum at a
cost of a large number of \texttt{tries}.

\begin{figure}
\centering
\includegraphics{paper_files/figure-latex/1d-2var-different-seeds-1.pdf}
\caption{The trace plot search better in a 1D projection problem with
two informative variables using different seeds (without polishing).
Since there are two informative variables, setting different value for
seed will lead search better to find either of the local maximum.}
\end{figure}

\begin{figure}
\centering
\includegraphics{paper_files/figure-latex/1d-2var-better-random-1.pdf}
\caption{Using search better random for the problem above will result in
finding the global maximum but much larger number of iteration is
needed.}
\end{figure}

@@

At previous two sections, only the iterator and the index value are
mapped onto the x and y aesthetics of the plot; while more aesthetics
i.e.~color, could be added to compare other parameters in the global
object. Two examples are shown below to explore and compare different
searching methods and neighbourhood parameter alpha.

\emph{Example: Polish} In principle, all the optimisation routines
should result in the same output on the same problem while this may not
be the case in real application. This motivates the creation of a
polishing search that polishes the ending basis and achieves unity on
different methods.

\texttt{search\_polish} takes the ending basis of a given search as the
current basis and uses a brutal-force approach to sample a large number
of basis (\texttt{n\_sample}) in the neighbourhood, whose radius is
controlled by \texttt{polish\_alpha}. Among the \texttt{n\_sample}
basis, the one with the largest index value becomes the candidate. If
its index value of the candidate basis is larger than that of the
current basis, it becomes the current basis in the next iteration. If no
basis is found to have larger index value than the current basis, the
searching neighbourhood will be shrunk and the search continues. The
polishing search ends when one of the four stopping criteria is
satisfied:

@@

@@

The following example conducted a 2D projection on the larger dataset
using search better with different configurations. \texttt{max.tries} is
a hyperparameter that controls the maximum number of try without
improvement and its default value is 25. As shown in Figure
\ref{fig:trace-compare}, after polishing, both trials attain the same
index value. However, a small \texttt{max.tries} of 25 is not sufficient
for the algorithm to find the true maximum. This is because 25 tries is
not sufficient for the 2D searching space.

@@

\emph{Example: The neighbourhood parameter alpha} Add an example on
comparing the neighbourhood parameter in search\_better \&
search\_posse.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# nrow(holes_2d_better_max_tries)}
\CommentTok{# nrow(holes_2d_pos)}
\end{Highlighting}
\end{Shaded}

@@

Sometimes, a parameter to explore could be a vector or matrix instead of
scalar, for example, in tour, we are interested to explore the
projection basis in the vector space. This imposes difficulties in
visualisation since we are bounded to perceive at most three dimensions.
Thus, principal component analysis is used to reduce the dimension of
projection bases and the first two principal components are mapped to
the x and y axis of the plot. Additional variable of interest could be
mapped to the color aesthetics to for exploration.

\emph{Example: understand search\_geodesic via mapping \texttt{info} to
color} \texttt{search\_geodesic} is a two-stage ascending algorithm with
four different stages in the search and a PCA plot useful to understand
how the algorithm works. Starting from the start basis, a directional
search is conducted in a narrow neighbourhood on five random directions.
The best one is picked and a line search is then run on the geodesic
direction to find the target basis. The starting and target basis are
then interpolated. In the next iteration, the target basis becomes the
current basis and then procedures continues. {[}should probably reword
this part with info levels xxx{]}

@@

\emph{Example: initial value for polishing alpha in search geodesic}
\texttt{search\_polish} is a brutal-force algorithm that evaluate 1000
points in the neighbourhood at each loop. Setting an appropriate initial
value for polish\_alpha would avoid wasting search on large vector space
that are not likely to produce higher index value. A default initial
value for polishing step is 0.5 and we are interested in whether this is
an appropriate initial value. The problem is a 1D projection of the
small dataset using \texttt{search\_geodesic} and followed by
\texttt{search\_polish}. Figure \ref{fig:polish-alpha} display all
projection bases on the first two principal components, colored by the
\texttt{polish\_alpha} parameter in the polishing step. Rather than
concentrating on the ending basis from \texttt{search\_geodesic} as what
polishing step is designed, it searches a much larger vector space that
are unnecessary. Thus a user-supplied smaller initial value for
\texttt{polish\_alpha} would be ideal.

On the right panel of Figure \ref{fig:polish-alpha-tracing}, the
\texttt{polish\_alpha} is initialised as the projection distance between
the last two target bases and we observe a concentrated searching space
near the ending basis. With the supplied initial value for
\texttt{polish\_alpha}, the algorithm reaches the same ending index
value with fewer iterations.

>>>>

\bigskip
\begin{abstract}

\end{abstract}

@@

Given an objective function \(f\), one way of optimising it is to equate
its gradient to zero. In modern optimisation problems, gradient
information can be hard to evaluate or sometimes even impossible and
Derivative-Free Optimisation (DFO) methods can be useful to approach
these problems. The two most common methods in DFO are
\emph{direct-search methods} and \emph{model-based methods} and this
paper dedicates to the discussion of direct-search methods, which gains
its popularity due to its simplicity in use and reliability in
complicated practical problem. Coined by \citet{hooke1961direct}, direct
search methods don't require any gradient or Hessian information and
evaluate \(f\) directly. A well-known example of it is the Nelder-Mead
algorithm \citep{nelder1965simplex} and it enjoys the popularity due to
its simplicity and reliability. {[}feel like this sentence can be
expanded to include more information. xxx{]}. Further, direct search
methods can be classified as \emph{stochastic} and \emph{deterministic}
depends on whether a random sample is used in the search. The stochastic
version of direct-search method is the focus of this paper.

@@

The optimisation problem we're interested in is in the context of
projection pursuit. Coined by \citet{friedman1974projection}, projection
pursuit is a method that detects the interesting structure
(i.e.~clustering, outliers and skewness) of multivariate data via
projecting it in lower dimensions. A Projection Pursuit Index (PPI) is a
function of the projection matrix (a.k.a projection angle, projection
basis) measuring the ``interestingness'' of data and we refer the value
of PPI function as \emph{index value}. In the literature, the indices
being proposed include Legendre index \citep{friedman1974projection},
hermite index \citep{hall1989polynomial}, natural hermite index
\citep{cook1993projection}, chi-square index
\citep{posse1995projection}, LDA index \citep{lee2005projection} and PDA
index \citep{lee2010projection}. {[}Any literature on holes index?
xxx{]}

@@

Projection pursuit is usually used in conjunction with a tour method
called \emph{guided tour}. Tour explores the multivariate data
\emph{interactively} via playing a series of projections, that form a
\emph{tour path}, like movie frames in real time. Starting from a
randomly generated projection basis, projection pursuit searches for a
target basis with higher index value and guided tour interpolates
geodesically between the initial basis and the target basis. This series
of bases forms a tour path and the projection of the data on these bases
is played in real time. This search-and-interpolate process is repeated
until no basis can be found with higher index value. Modified from
\citet{buja2005computational}, Figure \ref{tour-path} vividly depicts
the tour path in guided tour. Guided tour, along with other types of
tour, has been implemented in the \emph{tourr} package in R, available
on the Comprehensive R Archive Network at
\url{https://cran.r-project.org/web/packages/tourr/}
\citep{wickham2011tourrpackage}.

@@

\begin{itemize}
\item
  \emph{Being able to handle non-differentiable PPI function}: The PPI
  function could be non-differentiable, thus derivative free methods are
  preferred.
\item
  \emph{Being able to optimise with constraints}: The constraint comes
  from projection matrix being an orthonormal matrix.
\item
  \emph{Being able to find both local and global maximum}: Although the
  primary interest is to find the global maximum, local maximum could
  also reveal structures in the data that are of our interest.
\end{itemize}

@@

Random search methods has a black-box mechanism and focuses solely on
finding the global maximum point, while the projection pursuit problem
we have aims at \emph{exploring} the data and thus is interested in how
the algorithm finds its maximum. This motivates us \textbf{to develop a
visual diagnostic system for exploring the optimisation searching path}.

The necessity of developing such a system rather than simply producing
different diagnostic plots is because the diagnostics of each variable
requires a different function and these functions can't be scaled to
other problems. Thus we want to establish a set of rules that can
generalise the diagnostic of iterative algorithms.

The idea of generalising all the diagnostic plots under one framework is
inspired by the concept of \emph{grammar of
graphic}\citep{wickham2010layered}, which powers the primary graphical
system in R, ggplot2 \citep{ggplot2}. In grammar of graphic, plots are
not defined by its appearance (i.e.~boxplot, histogram, scatter plot
etc) but by ``stacked layers''. By this design, ggplot doesn't need to
develop a gazillion of functions that each produces a different type of
plot. Instead, it creates aesthetic mapping from the variables to the
geometric objects and builds the plot through different layers.

{[}Do I need this paragraph to details the grammar of graphic? xxx{]}

\hypertarget{global-object}{%
\subsection{Global object}\label{global-object}}

Ggplot requires a data frame that contains all the variables to plot and
a \emph{global object} is constructed as the data frame supplied to the
visual diagnostic plots to better suit the characters of iterative
optimisation algorithms. Given an optimisation algorithm, two primary
variables of interest are the \emph{decision variable} and the
\emph{value of the objective function}. \emph{iterators} indexes the
data collected and this order has time series feature that prescribes
the progression of the optimisation. Given the complexity of the
algorithm, there could be multiple iterators. There are other parameters
that can't be classified as one of the three categories but are also of
our interest. They are defined under the fourth category: \emph{other
parameter of interest}. This could include variables that are specific
for one particular algorithm. Below we show an example of defining the
global object of projection pursuit guided tour.

In the projection pursuit guided tour, the decision variable is the
projection matrix, the value of objective function is the index value.
There are three iterators: \texttt{id} is the smallest ordering unit
that increases by one for each observation; \texttt{tries} is updated
once a search-and-interpolate step is finished. A third iterator
\texttt{loop} is used to record the number of repetition in the search
step and starts over from one at the beginning of a new \texttt{tries}.
Three other parameters of our interest includes \texttt{method},
\texttt{alpha} and \texttt{info}. \texttt{method} identifies the name of
the searching method used and we are interested in comparing the
performance between different algorithms under direct search. The
neighbourhood parameter \texttt{alpha} controls the size of the sampling
space and we are interested to understand how the searching space
shrinks as the algorithm progresses. \texttt{info} labels different
stages in the searching process. A sketch of the global object for
projection pursuit guided tour is presented in Figure \ref{fig:glb-obj}.

@@

Two set of simulated data are used in the demo of the visual diagnostics
in projection pursuit guided tour. A small dataset consists of 1000
randomly simulated observations of five variables (\texttt{x1},
\texttt{x2}, \texttt{x8}, \texttt{x9}, \texttt{x10}). \texttt{x2} is the
informative variable simulated from two bi-modal normal distribution
centred at -3 and 3 with variance of one and the other four are
simulated from non-informative standard random normal distributions. The
data has been scaled to ensure \texttt{x2} has variance of 1. The goal
is to find the 1D projection of bi-modal shape, which corresponds to the
projection matrix (vector) of \texttt{{[}0,\ 1,\ 0,\ 0,\ 0{]}}.

A larger dataset contains more informative variables (\texttt{x3} to
\texttt{x7}) of different types. The distribution of all the variables
except \texttt{x3} is plotted in Figure \ref{origin-data} and
\texttt{x3} takes 500 positive one and 500 negative one. {[}should I
introduce the dist for each var? xxx{]} {[}also feel like we don't
really use \texttt{x3} to \texttt{x6}, should we not mention about
these? xxx{]}

\begin{figure}
\centering
\includegraphics{paper_files/figure-latex/origin-data-1.pdf}
\caption{\label{origin-data} The distribution of simulated data except
x3}
\end{figure}

In tour, when the optimisation ends the global object will be stored and
printed and can be turned off via \texttt{print\ =\ FALSE}. Additional
messages during the optimisation can be displayed via
\texttt{verbose\ =\ TRUE}. Below presented the global object of the 1D
projection using the small dataset. {[}the printing is not ideal as it
doesn't show all the columns. xxx{]}

@@

@@

In tour, all the points recorded in the global object can be divided
into two broad categories: searching points and interpolating points

\begin{itemize}
\item
  \emph{Searching points} include the observations recorded in the
  searching algorithm to find the target basis. The points for target
  bases is also included in the searching points and there is one such
  point per \texttt{tries}.
\item
  \emph{interpolating points} exist in the guided tour to produce
  continuous animated view from one target basis to another and it
  doesn't have \texttt{loop} value.
\end{itemize}

@@

The largest difficulties of exploring searching points is its unknown
number of observations per \texttt{tries}. Mapping \texttt{id} on the
x-axis will leave the \texttt{tries} with few observations a small space
in the plot, while those \texttt{tries} with large number of search
points (usually towards the end) occupy the vast majority of the plot
space. This motivates to summarise the points in each \texttt{tries}. At
each iteration, rather than knowing the index value of \emph{every}
points, we are more interested to know a general summary of all the
points and more importantly, the point with the largest
\texttt{index\_val} since it prescribes the geodesic interpolation and
future searches.

Boxplot is a suitable candidate that provides five points summary of the
data, however it has one drawback: it doesn't report the number of point
in each box. We may risk losing information on how many points it takes
to find the target basis by displaying the boxplot alone for all
\texttt{tries}. Thus, the number of point for each \texttt{tries} is
displayed at the bottom of each box and we provide options to switch
\texttt{tries} with small number of points to a point geometry, which is
achieved via the \texttt{cutoff} argument. A line geometry is also added
to link the points with the largest index value in each \texttt{tries}.
This helps to visualise the improvement made by each \texttt{tries}.

\emph{Example: exploring searching points} The larger dataset is used
for 2D projection with \texttt{search\_better} and
\texttt{max.tries\ =\ 500}. In Figure \ref{points}, the searching points
with \texttt{id} and \texttt{tries} on the x-axis, colored by
\texttt{tries}. Label at the bottom indicates the number of observations
in each tries and facilitates the choice of cutoff to switch from point
geometry to boxplot geometry (\texttt{cutoff\ =\ 15}). The line geometry
suggests the largest improvement happens at \texttt{tries\ =\ 5}.

@@

@@

Plotting the interpolating points as time series data allows us to
diagnose characteristics of different configurations and index
functions. Here we present two examples of using plots to diagnose tour
algorithms and different index functions.

\emph{Example: Interruption} This examples uses \texttt{search\_better}
for a 2D projection on the larger dataset using the \texttt{holes}
index. In the interpolation stage, continuous frames will be constructed
to bridge the project on the current basis to the target basis. The
target basis will become the current basis in the next iteration and the
searching stage continues to find the next target basis. In the left
panel of Figure \ref{interruption}, we observe that there are
interpolating bases with higher index values than the target basis and
these bases could be used to search for new basis in the next searching
stage.

An interruption is then constructed to accept the interpolating bases up
to the one with the largest index value and use that basis as the
current basis for the next searching stage. After implementing this
interruption, the tracing plot with the same configuration is shown on
the right panel of Figure \ref{interruption}. In the third
\texttt{tries}, rather than interpolating fully to the target basis at
\texttt{id\ =\ 62}, it stops before the index value starts to decrease
at \texttt{id\ =\ 60}. This implementation results in a higher index
value in the end with fewer steps.

@@

\emph{Example: Noisy index function} The interpolation path of holes
index, as seen in Figure \ref{interruption}, is smooth, while this may
not be the case for more complicated index functions. \texttt{kol\_cdf}
index, an index function based on Kolmogorov test, compares the
difference between a projection matrix and a randomly generated normal
distribution based on cumulated distribution function (CDF). Several
diagnostic visualisations below show the characteristic of this index
function.

Figure \ref{kol-cdf} compares the tracing plot of the interpolating
points for \texttt{search\_geodesic} and \texttt{search\_better} on 1D
projection for \texttt{kol\_cdf} index and the interpolation path shows
a zig-zag pattern. Polishing step has done much more work to reach the
final index value for \texttt{search\_geodesic} than
\texttt{search\_better} and this indicates that noisy index's favour of
a random search method than ascent method {[}the order of the examples
needs to be rearranged since I haven't introduced polishing here.
xxx{]}.

\begin{figure}
\centering
\includegraphics{paper_files/figure-latex/kol-cdf-1.pdf}
\caption{\label{kol-cdf}Comparison of two different searching methods:
search\_geodesic and search\_better on 1D projection problem for a
noisier index: kol\_cdf. The geodesic search rely heavily on the
polishing step to find the final index value while search better works
well.}
\end{figure}

The second example for \texttt{kol\_cdf} index uses 1D projection on the
larger dataset. Since there are two informative variables and the
projection is 1D, there are two local maximum when the projection matrix
at \([0, 1, 0, 0, 0, 0]\) and \([0, 0, 1 ,0, 0, 0]\). As in Figure
\ref{1d-2var-different-seeds}, different local maximum can be found
using \texttt{search\_better} with different seeds while
\texttt{search\_better\_random} can always find the global maximum, as
in Figure \ref{1d-2var-better-random}, although at a high cost of number
of points evaluated.

\begin{figure}
\centering
\includegraphics{paper_files/figure-latex/1d-2var-different-seeds-1.pdf}
\caption{\label{1d-2var-different-seeds}The trace plot search better in
a 1D projection problem with two informative variables using different
seeds (without polishing). Since there are two informative variables,
setting different value for seed will lead search better to find either
of the local maximum.}
\end{figure}

\begin{figure}
\centering
\includegraphics{paper_files/figure-latex/1d-2var-better-random-1.pdf}
\caption{\label{1d-2var-better-random}Using search better random for the
problem above will result in finding the global maximum but much larger
number of iteration is needed.}
\end{figure}

@@

\emph{Example: Polish} In principle, all the optimisation routines
should result in the same output on the same problem while this may not
be true in real application. This motivates the creation of a polishing
search that polishes the ending basis and achieves unity on different
methods.

\texttt{search\_polish} takes the ending basis of a given search as the
current basis and uses a brutal-force approach to sample a large number
of basis (\texttt{n\_sample}) in the neighbourhood, whose radius is
controlled by \texttt{polish\_alpha}. Among the \texttt{n\_sample}
basis, the one with the largest index value becomes the candidate basis.
If its index value of the candidate basis is larger than that of the
current basis, it becomes the current basis in the next iteration. If no
basis is found to have larger index value than the current basis, the
searching neighbourhood will be shrunk and the search continues. The
polishing search ends when one of the four stopping criteria is
satisfied:

@@

{[}should the stopping criteria be more detailed? xxx{]}

@@

The following example conducted a 2D projection on the larger dataset
using search better with different configurations. \texttt{max.tries} is
a hyperparameter that controls the maximum number of try without
improvement and its default value is 25. As shown in Figure
\ref{trace-compare}, both trials attain the same index value after
polishing while a small \texttt{max.tries} of 25 is not sufficient for
\texttt{search\_better} to find the global maximum. The
\texttt{max.tries} argument needs to be adjusted to ensure it is
sufficient to explore the parameter space.

@@

@@

The parameter to explore can be a vector or matrix instead of scalar,
for example, the projection basis in tour. This imposes difficulties in
visualisation since humans are bounded to perceive at most three
dimensions in a plot. Thus, principal component analysis is used to
reduce the dimension of projection bases and the first two principal
components are mapped to the x and y axis of the plot. Additional
variable of interest could be mapped to the color aesthetics for
exploration.

\emph{Example: understand search\_geodesic} {[}feel like this example is
merely explaining search geodesic algorithm, so maybe introduce the
animated plot here? xxx{]} \texttt{search\_geodesic} is a two-stage
ascending algorithm with four different stages in the search and a PCA
plot useful to understand how the algorithm progresses and the relative
position of each basis in the PCA projected 2D space. Starting from the
start basis, a directional search is conducted in a narrow neighbourhood
on five random directions. The best one is picked and a line search is
then run on the geodesic direction to find the target basis. The
starting and target bases are then interpolated. In the next iteration,
the target basis becomes the current basis and then procedures
continues.

@@

\emph{Example: initial value for polishing alpha}
\texttt{search\_polish} is a brutal-force algorithm that evaluate 1000
points in the neighbourhood at each loop. Setting an appropriate initial
value for polish\_alpha would avoid wasting search on large vector space
that are not likely to produce higher index value. The default initial
value for polishing step is 0.5 and we are interested in whether this is
an appropriate initial value to use after \texttt{search\_geodesic}. The
problem is a 1D projection of the small dataset using
\texttt{search\_geodesic} and followed by \texttt{search\_polish}. The
top-left panel of Figure \ref{polish-alpha} displays all the projection
bases on the first two principal components, colored by the
\texttt{polish\_alpha}. We can observe that rather than concentrating on
the ending basis from \texttt{search\_geodesic} as what polishing step
is designed, \texttt{search\_polish} searches a much larger vector
space, which is unnecessary. Thus a customised smaller initial value for
\texttt{polish\_alpha} would be ideal. One way to do this is to
initialised \texttt{polish\_alpha} as the projection distance between
the last two target bases. The top-right panel of Figure
\ref{polish-alpha} shows a more desirable concentrated searching space
near the ending basis. Both specifications of initial value allow the
searches to reach the same ending index values.

\begin{figure}
\centering
\includegraphics{paper_files/figure-latex/polish-alpha-1.pdf}
\caption{\label{polish-alpha}PCA plot of two different polish alpha
initialisations. A default polish alpha = 0.5 searches a larger space
that is unncessary while a small customised initial value of polish
alpha will search near the ending basis. Both intialisations reach the
same ending index values.}
\end{figure}
